---
bibliography: ../../references.bib
title: "Distinct functional responses of consumers and their producers to climate drive mutualistic  network asymmetry"
execute: 
  eval: true
format:
  html:
    toc: true
    number-sections: true
editor: visual
abstract: |
  Functional traits are often used to infer the ecological processes that determine the composition of species assemblages. Whereas most trait-based approaches to infer community assembly processes focus on a single trophic level, traits also mediate interactions between trophic levels. Owing to the matching of traits facilitating interactions between producer and consumer assemblages, the functional trait diversity of different trophic levels is expected to covary in space. However, the differential response of consumers and producers to environmental gradients can cause a decoupling of functional diversity between trophic levels, which we coin functional trophic asymmetry (FTA). Here, we develop a framework to quantify FTA with incomplete trait and interaction data. We use this framework to infer the processes underpinning multitrophic community assembly and explore their role in shaping the topological structure of ecological networks. We used digitally available data on functional traits, pairwise mutualistic interactions, and geographic distributions of consumers (mammalian frugivores) and their producers (palms) to quantify FTA for assemblages occurring in the Neotropics. To cover major data gaps between species-level traits and interaction data at finer spatial grain, we trained machine learning models to downscale the Neotropical metaweb to grid cell-level networks. For each grid-cell, we estimated FTA for all combinations of interaction guilds, defined as distinct subsets of producer and consumer assemblages sharing partners and playing similar roles within mutualistic networks. We used generalized additive models to relate geographic variation in FTA to climatic variables and assessed whether the strength of these relationships varied among guilds. Finally, we examined the relationship between FTA and network specialization. Assemblage-wide FTA was positively related to mean annual temperature across the Neotropics. For individual guild combinations, the positive relationship between FTA and precipitation seasonality was greater for those predicted to interact strongly. Network specialization was positively related to FTA across all combinations. These results suggest that future increases in temperature and precipitation seasonality may lead to more specialized mutualistic networks that are more vulnerable to collapse when facing additional threats and local extinctions.
---

```{r}
#| message: false
#| warning: false
#| include: false

library(tidyverse)
library(gt)

```

## Introduction

Ecologists often examine patterns of functional trait diversity to investigate community assembly processes [@ackerly2003community; @kraft2015community]. To date, however, trait-based approaches in ecology often focus on a single trophic level, whereas approaches that consider multiple trophic levels remain rare [@lavorel2013novel; @seibold2018necessity]. An approach that considers processes operating within and between trophic levels is necessary to better understand the assembly of multitrophic communities [@allesina2008general; @marjakangas2022trait; @saravia2022ecological]. Moreover, considering trophic interactions while studying community assembly could shed new light on processes underpinning ecological networks [@allesina2008general]

Classical approaches to study community assembly rely on the concept of environmental filtering, sorting or selection, where density independent conditions constrain the functional richness of species assemblages [@laliberte2010distance; @villeger2008new; @hillerislambers2012rethinking; @kraft2015community]. Functional richness refers to the variability and relative frequency of different functional traits observed in a community. It is often used to estimate the strength of selection imposed by the environment [@kraft2008functional; @kraft2010functional; @kraft2015community]. High functional richness can indicate weak environmental selection whereas low functional richness can indicate strong selection [@kraft2008functional; @halpern2008functional; @paine2011functional]. In a multitrophic context, the effects of environmental selection can cascade to other trophic levels such that selection on consumer traits can shape the functional richness of their resources, modulated by their degree of reciprocal dependency or co-evolution [@lavorel2013novel; @guzman2019towards]. Moreover, the same environmental gradient could exert selective pressures of different strength on communities at distinct trophic levels [@marjakangas2022trait]. Differences in the strength of selective pressure among trophic levels could then possibly constrain the structure or topologies of trophic networks [@bluthgen2007specialization; @schleuning2012specialization; @dehling2014functional; @dehling2020similar].

Inferring the relative strength of environmental selection between trophic levels requires high-dimensional approaches to deal with sparse observations for many species [@rohr2010modeling; @strydom2022food]. We introduce the concept of functional trophic asymmetry (FTA), which allows inferring the relative influence of environmental selection and trait matching on the composition of multitrophic assemblages (Figure 1). FTA is the difference in the richness of interaction-relevant traits between trophic levels in a multitrophic network. FTA can occur because traits mediating species interactions (i.e., interaction niches) across trophic levels can also mediate the responses of species to their abiotic environment (i.e., environmental niches) [@mccain2014body; @moretti2009combining; @dehling2020similar]. As an example, plant seed size determines the outcome of animal-mediated seed dispersal [@donoso2017complementary; @donoso2020downsizing] as well as physiological limits, such as tolerances of plant seedlings to desiccation [@hoekstra2001mechanisms]. High FTA could indicate differences in the strength of environmental selection over the interaction niches of distinct trophic levels within a multitrophic species assemblage. Alternatively, low FTA could indicate that the strength of the environment selection shaping interaction niches is similar between trophic levels, e.g., equally weak or equally strong [@marjakangas2022trait]. When interactions between producers and consumers are mutualistic, low FTA could also emerge under strong trait matching and therefore indicate the influence of trait co-evolution during multitrophic community assembly [@dehling2014functional ; @albrecht2018]. By studying spatial variation in FTA along environmental gradients, we sought for the conditions promoting environmentally versus cross-trophic interaction-driven community assembly [@schleuning2020trait; @bello2023analyzing].

![**Figure 1: In ecological networks, bottom-up assembly refers to network structure being mainly shaped by the resource community, while top-down assembly reflects stronger influence from the consumer community.** Although both processes may act together, their relative influence likely shifts along environmental gradients. This conceptual model illustrates the relationship between functional richness (FR) and functional trophic asymmetry (FTA) along environmental gradients and implications for network specialization. (a) Left: variation in FR for producers (depicted in green) and consumers (depicted in orange) along an environmental gradient. As environmental conditions become harsher along climatic gradients, it can result in a decline in functional richness due to stronger filtering of community composition. This decline can occur at different rates, leading to two scenarios: Top panel: differential decline in FR. If consumer FR declines more strongly than producer FR, it can indicate that the mode of network assembly is top-down. If producer FR declines more strongly than consumer FR, it can indicate that network assembly is bottom-up. This results in a substantial increase in FTA. Bottom panel: parallel decline in FR reflects both producer and consumer FRs reducing at a similar rate. FTA then remains relatively constant along the gradient. This scenario indicates a balanced impact of environmental changes across trophic levels, preserving the relative functional relationship between producers and consumers. (b) Network specialization indices can measure how generalized the interactions are between producers and consumers within ecological networks, revealing the role of changing FTA for network specialization.](images/clipboard-4279899070.png)

Frameworks linking multitrophic functional diversity to network topology along broad-scale environmental gradients are crucial to understand the effects of global change on biodiversity and ecosystem function [@bello2023analyzing; @dehling2018structure; @schleuning2020trait; @schleuning2014ecological; @schleuning2023animal]. Functional responses of consumer and producer assemblages to climate influence functional richness at the level of the multitrophic community [@garcia2018frugivore]. Because some of these traits are involved in interactions across trophic levels, the filtering of traits along environmental gradients could constrain the identity, number, and frequency of species interactions, and therefore network topology [@marjakangas2022trait; @emer2020seed; @albrecht2018]. As an example, constraints of varying intensities along climatic gradients, which limit the relative availability of interaction partners across trophic levels, could influence emergent patterns of network structure such as the specialization of multispecies interactions [@bluthgen2007specialization; @bluthgen2006measuring; @marjakangas2022trait]. While high levels of network specialization represent networks predominantly made of “one-to-one” interactions, low levels of network specialization represent networks with species showing predominantly “one-to-many” interactions [@bluthgen2006measuring; @bluthgen2011functional] (Figure 1B). One highly expected outcome is that when functional trophic asymmetry is high, networks will have low specialization. For example, take a plant community exhibiting a low richness of flower displays associated with a bee community (pollinators) exhibiting a wide variety of proboscis lengths. These plants are unlikely to form “one-to-one” interactions with only a subset of bees having matching proboscis length. Otherwise, non-matching pollinators would have no food resources and be extirpated. By partitioning deviations from expected FTA and network specialization relationships with null models, one can separate the relative influences of processes operating between trophic levels (e.g. trait matching) from those within trophic levels (e.g. environmental selection) in network assembly [@marjakangas2022trait]. Despite this potential, the relationship between network specialization and functional trophic asymmetry has not been comprehensively explored.

Preserving mutualistic interactions between palms and their mammalian frugivores is important to sustain biodiversity and ecosystem function in the tropics [@bogoni2020extent; @marques2022mutualism]. Mammalian frugivores facilitate the dispersal of palm fruits, which helps to prevent local extinctions amid disturbance and to maintain biodiversity in these ecological networks [@acevedo2020sampling; @messeder2020searching]. To effectively preserve these interactions, it is crucial to understand how co-occurring palm (producer) and mammalian frugivore (consumer) communities respond to environmental gradients. By examining co-variation in their functional richness across broad geographic scales and linking those patterns to spatial and/or temporal variation in climate (Figure 2), we can identify key abiotic factors that influence the assembly of their mutualistic relationships. Here, we ask **(1)** which climatic variable(s) best explain(s) geographic variation in the functional richness of palms and mammalian frugivores, **(2)** whether differences in these relationships lead to FTA, and **(3)** which climatic variable best explains geographic variation in FTA across the Neotropics. We also ask **(4)** whether the strength of interactions between palm-frugivore interaction guilds relates to the strength of the relationship between FTA and climate, and **(5)** whether geographic variation in FTA relates to those of network specialization.

![**Figure 2: Workflow for integrating ecological, trait, and distributional data to model species interactions and infer ecological networks across geographic space.** Digitally available information—including interaction records, trait measurements, and expert-based range maps—originate from literature, museum specimens, and field collections. These data streams are used in two main ways: (1) network parametrization using latent trait models (left panel), which estimate interaction probabilities (Pᵢⱼ) between palm and mammal species based on inferred compatibility in unobserved trait dimensions; and (2) trait linkage, where observed traits are statistically associated with latent traits to improve biological interpretability. Independently, species distribution data are used to assemble multitrophic communities (palm–mammal co-occurrence) within spatial grid cells (e.g. across the Neotropics). These grid-based assemblages are then combined with the interaction probability matrix (Pᵢⱼ) to infer local (grid cell) ecological networks. The outputs include palm–mammal interaction matrices (green–orange bipartite graphs) and derived network properties. The letters LP and POC represent two components of the generative model: Latent Probabilistic (LP) and Probability of Cooccurrence (POC), respectively, reflecting alternative formulations for generating spatialized interaction predictions.](images/clipboard-3327170127.png)

## Methods

### Study system

We focused on multitrophic assemblages of Neotropical palms and their mutualistic, seed dispersing, mammalian frugivores. Palms (Plantae: Arecaceae) are a keystone plant family from tropical regions that provides fruit resources to a wide variety of vertebrate frugivores, including birds and mammals [@munoz2019synthesis; @zona1989review]. Frugivorous mammals (Animalia: Mammalia) are among the most important palm-seed dispersers, particularly over long distances. Most frugivorous mammals feeding on palms are seed eaters and pulp eaters, dispersing palm seeds mostly via ectozoochorus dispersal [@messeder2020searching; @marques2022mutualism]. Frugivory-related traits have notably underlain palm diversification and played a key role in the evolution of palm traits [@onstein2014diversification; @onstein2017frugivory]

## Geographic distribution data

We obtained binary species distribution data (presence/absence) of palms from the geographic range maps of [@bjorholm2005environmental] and on mammals from the IUCN (International Union for the Conservation of Nature) data portal. To generate local gridded multitrophic species assemblages across the Neotropics, we intersected the species-level range maps with a spatial grid where each grid cell represented a 1 $\times$ 1 degree latitude and longitude resolution across the extent of the entire Neotropics. We then recorded all palm and mammal frugivore species occurring in each grid cell as our grid-cell level multitrophic assemblages.

### Trait data

We collected species-level multitrophic trait data related to the physiological tolerance of palms and frugivorous mammals to the abiotic environment and to their mutualistic interactions. For palms, we extracted data from the PalmTraits 1.0 dataset [@kissling2019palmtraits]. We collected data on growth form, maximum stem height, and average fruit length. For frugivorous mammals, we obtained trait data from the EltonTraits 1.0 database [@wilman2014eltontraits]. We selected data on body mass, diet, and daily activities. Diet data from the EltonTraits 1.0 database is coded as percentage use distribution across ten diet categories. We excluded from our analysis species without fruit in their diet. Activity was coded as a dummy variable with three categories (Diurnal, Crepuscular, Nocturnal). Finally, body mass was coded as a numerical variable in kg. We excluded bats from the analysis as almost no Neotropical bat species is feeding on palm fruits [@messeder2020searching]. From this dataset, we selected only those species whose geographic range in the 1×1° latitude grid co-occurred with at least 5 other palm and mammal frugivore species. In total, we worked with a set of 494 palm species and 488 mammal frugivore species with linked trait and geographic data, which comprised respectively \~61% [@roncal2013cenozoic; @balslev2016palms] and \~64% [@fuzessy2025neofrugivory] of reported species for the Neotropics .

### Pairwise interaction data

We used data on seed dispersal interactions between palms and frugivorous mammals for the Neotropics, originating from recollections of seed dispersal records found in the published literature and interaction records recorded at the species level [@munoz2019synthesis]. Each pairwise species interaction record reflects where an article mentions the fruit or the seed of a palm being dispersed, carried, or defecated by a frugivorous mammal. Interaction records collected in this database were previously vetted by [@munoz2019synthesis] to reflect effective seed dispersal interactions, while avoiding those that reflect mere seed consumption (vetting criteria found in: [@munoz2019synthesis]). In total, we gathered a total of 581 interaction records between 69 palms and 111 frugivore mammals.

### Climatic data

We used bioclimatic variables from WorldClim [@fick2017worldclim] to represent large-scale spatial and temporal variation of climate in the Neotropics. Specifically, we used mean annual temperature (BIO01), total annual precipitation (BIO12), temperature seasonality (BIO04), and precipitation seasonality (BIO15). We selected these variables to represent the main axes of environmental niche variation shared by palms and frugivorous mammals (representing 91% of the total variance contained in the 19 WorldClim BIOCLIM variables). To harmonize spatial resolution with the scale of ecological analyses, each selected raster layer was aggregated by averaging approximately 34 neighboring cells into one while preserving broad-scale climatic gradients relevant to the study (i.e. 1 $\times$ 1 degree grid resolution).

### Statistical analysis

#### Building a probabilistic Neotropical metaweb from aggregated binary interaction records

We fitted latent variable network structural models [@rohr2010modeling; @terry2020finding] based on different underlying assumptions to infer interaction probabilities from binary observations of species interactions (Figure 3). Specifically, we tested the stochastic block model, the connectance model, the trait-matching model, and the matching-centrality model [@terry2020finding]. The stochastic block model assumes that ecological networks are modular, with species of consumers interacting more within their preferred groups of producers (i.e., interaction guilds) than outside their preferred groups. This model outputs three incidence matrices, reflecting predicted interactions: (i) one with guilds of palm species based on their modular interactions with mammals, (ii) a similar one for mammals, and (iii) one representing the interaction probabilities (Theta) among the guilds of each group (i.e., palm guilds and mammal guilds, respectively). The connectance model posits that interactions of specialist species are subsets of those of generalist species and optimizes connectivity scores, constrained by this assumption, to recreate observed network patterns. The trait-matching model assumes non-random species interactions determined by trait differences, optimizing parameters along latent-trait axes. Finally, the matching-centrality model combines connectivity scores and latent-trait axes [@terry2020finding]. Additional details about the model assumptions are explained in Supplementary B1. We fitted these models to our available interaction data and selected the model that best predicted the observed Neotropical pattern of seed dispersal interactions. Using Youden’s J as a metric that balanced model sensitivity and specificity [@poisot2023guidelines], we found that the stochastic block model was the best supported model (Figure 3) and therefore focus on it in the rest of the manuscript.

![**Figure 3: Evaluation of four network assembly models to predict the binary structure of observed palm-mammal frugivory interactions in the Neotropics.** (a) Region under the Curve (ROC) curves, (b,c) Youden's J index, and (d) clustering heatmaps. The four models are the Centrality Model (CM), Trait Matching Model (TMM), Matching Centrality Model (MCM), and Stochastic Block Model (SBM). The ROC curves in (a) assess model performance through the relationship between model sensitivity (i.e. true positive rate) and model specificity (true negative rate). The curves above the diagonal demonstrate stronger predictive ability, showing that the model performs significantly better than random guesses in identifying ecological interactions. If a ROC curve is close to or below this diagonal, the model's predictive performance is no better than random. The boxplot in (b) summarizes the Youden's J index for each network assembly model, where higher values reflect higher overall predictive accuracy. In (c), the Youden’s J index shows variation over different probability thresholds to specify binary interactions. In (d), heatmaps highlight structural differences among the four models.](images/clipboard-387333362.png)

#### Identifying interactions guilds

The hyperparameters of the Stochastic Block Model provided the best fit for capturing the observed interactions, suggesting that palm - frugivorous mammal interaction networks are highly modular. Certain groups of producers were thus more likely to interact with certain groups of consumers than others, and vice versa. In this context, we define an interaction guild as a distinct group of palm and frugivorous mammal species within the Neotropical metaweb that exhibits similar interaction patterns, forming a module. Within each guild, both producer and resource species perform comparable functional roles in the network. Within and between each guild combination of consumers and producers, species pairs exhibit the same interaction strength, where interaction strength is defined as the magnitude or intensity of the effect that one species has on another within an ecological network. Here we estimate interaction strength as the probability that a species pair would interact in nature. The stochastic block model estimates such interaction probabilities based on the frequency of interactions observed between species assigned to a given guild of consumer or producer. The stochastic block model uses maximum likelihood to adjust the number of guilds and the distribution of interaction probabilities within and between guilds such that they best explain the observed pattern of interactions. Using stochastic block models largely reduces the complexity of dealing with pairwise interaction strengths by treating them as a guild-level phenomenon instead of a species-specific one. Importantly, there is no restriction for the stochastic model to assign palm or mammal species to every guild; consequently, some guilds may contain zero species within the Neotropical metaweb or a given grid cell. Our approach resulted in a theta matrix representing 7 x 7 latent guild combinations. In addition, we obtained a guild-association matrix for each trophic level. For frugivorous mammals, five guilds (guilds 1, 3, 4, 6, and 7) had at least one mammal species and two inferred guilds with no mammal species allocated to them (a.k.a empty guilds: 2, 5). For palms, we obtained four guilds (guilds 2, 3, 5, and 6) allocating at least one palm species and three empty guilds (1,3,7).

#### Downscaling the Neotropical metaweb to generate grid-cell level networks

The availability of primary biodiversity data on palms and their frugivorous mammals was imbalanced, with a high availability of distribution ranges and species traits, but a limited number of interaction records. We therefore used a two-fold approach to downscale the initial, Neotropical metaweb to include interactions between every potentially co-occurring palm and mammal frugivore in every grid cell across the Neotropics.

First, we employed multinomial logistic regression models using neural networks to predict the interaction guild affiliation of each species from the stochastic block model results using species-level trait data. We justify the choice of multinomial logistic regression models as these can handle the prediction of non-binary outcomes, such as the labeling of interaction guilds per species. Separate multinomial logistic regression models using a single-hidden layer network were fitted for palms and frugivorous mammals, respectively [@ripley2016package] with 75% of the data allocated for training and the remaining 25% for testing. This approach is useful when dealing with multicollinearity, as it can learn complex and non-linear relationships and interactions among multiple predictor variables. This allowed us to separate the relative importance of distinct matching traits from the stochastic block model group affiliations. We extracted variable importance scores based on combinations of the absolute values of the best fit model weights [@gevrey2003review].

Second, we considered local pairwise species interaction probabilities as the product of the values from the Theta matrix from the stochastic block model that represents the latent interaction probabilities between species pairs within and between guilds multiplied by their probability of co-occurrence in a grid cell. To represent species’ co-occurrence probabilities, we calculated a spatial overlap index for each species pair as the product of their relative range areas, normalized by the sum of their areas. Additionally, we computed the geographic distance between species' range centroids and rescaled these distances to a 0–1 scale, with proximity increasing interaction weight. The final interaction weight for each species pair was computed as the product of the rescaled SBM-derived probability, the rescaled spatial overlap index, and the inverse of the rescaled geographic distance. In other words, grid-level interaction probabilities represent the biological feasibility of a pairwise interaction given the regional availability of their traits [@strydom2022food], with the overall assumption of higher species abundances toward the center of their geographic ranges [@schmidt2022population]

#### Estimating functional richness

Functional richness (FR) is typically defined as the variability and relative frequency of functional traits within a community [@kraft2008functional ; @kraft2010functional]. The interaction guilds identified by the stochastic block model encapsulate unique combinations of functional traits reflected in species interactions. Consequently, the guild affiliation itself represents a categorical multitrophic “functional trait syndrome” for our analysis. For each grid cell, we calculated the proportion of palm and mammal species belonging to each guild (n = 7) relative to the total number of palm or mammal species present locally. This normalization ensures that our measure reflects trait diversity rather than simply species richness. Thus, our metric of FR quantifies the diversity of functional roles occupied by palms and mammals within local networks, aligning conceptually with broader ecological definitions of FR but adapted to multitrophic assemblages through the latent trait representation provided by the stochastic block model. We assigned a value of FR = 0 to those guilds with no palm or mammal species present.

#### Estimating functional trophic asymmetry

We quantified FTA as the absolute difference in functional richness between palm and frugivorous mammal assemblages within each of the 1,242 grid cells. The FTA measure for each grid cell was obtained by computing the absolute difference between guild-specific functional richness values across the two trophic levels, resulting in one FTA value per guild combination per grid cell. This approach resulted in 43 unique realized pairwise guild combinations, after excluding 6 combinations where both guilds lacked species representation in the Neotropical metaweb. Overall, 98% of all available guild combinations exhibited nonzero FTA values across grid cells, indicating that only 2% of potential palm–mammal guild interactions from the metaweb were locally absent. To estimate a single grid-cell level metric of functional trophic asymmetry, we summed the FTA values across all interaction guilds, weighted by their respective interaction strengths (FTAsum). This approach was chosen as it accounts for the uneven contributions of each guild to network structure.

#### Estimating network specialization

We estimated network specialization for each grid cell using the metric H2’. H2’ is a network-level index that varies between 0 and 1 (Blüthgen et al. 2007). High values indicate networks that are more specialized, meaning that species from one trophic level interact with only one or just a few species in the other trophic level. Low H2’ values indicate that there is a low specificity of interactions in the network, meaning that species from one trophic level interact with multiple species of the other trophic level. Because inferred networks varied in their network size (i.e., number of unique interactions between palms and frugivorous mammals), we rarefied the H2’ computed for each grid-cell level network such that they would all have the same size (i.e., number of interactions). Specifically, we rarefied all networks to 100 pairwise interactions and repeated the procedure 999 times to get a distribution of rarefied H2’ values (Terry and Lewis 2020). We then selected the median of this H2’ distribution as our grid cell-level measure of network specialization.

#### Assessing the influence of climate on FTA

To assess whether climate has an influence on FTA, we fitted a Generalized Additive Model (GAM) to examine the relationships between FTA and the four bioclimatic predictors described above. The GAM approach allows for modelling flexible non-linear relationships between the predictors and the response variable using smoothed penalized functions (Wood 2017). We fitted separate splines for each climatic predictor.

#### Assessing how interaction strength mediates the influence of climate on FTA

In our approach, interaction strength represents the estimated frequency of interactions between species assigned to a given guild of consumer or producer, that is, the values from the theta matrix from the SBM (Figure S1). To assess whether the strength of interactions between producer and consumer guilds mediates the relationship between FTA and climate, we included “interaction strength” as an interaction term in the GAM, allowing splines between FTA and climate to vary non-linearly depending on the strength of interaction among guilds.

#### Assessing the relationship between FTA and H2’

We used a GAM to investigate the relationship between rarefied grid-cell network specialization (H2’) as a response variable and FTA (z-scores) as the main predictor. We also added covariate functions to account for the effect of mean annual temperature, total annual precipitation, precipitation seasonality, and temperature seasonality because these climate variables may influence H2’ independently of FTA. This allowed us to isolate the specific impact of FTA on H2’ while controlling for the indirect effects of climate on FTA.

## Results

### Interaction guild delineation

The result of our modelling approach was one theta matrix reflecting the modular pattern of interaction guilds in the Neotropical metaweb (Figure S1) and two association matrices linking frugivorous mammals (Figure S2a) and palms (Figure S2b) to their interaction guilds. Together, they represent high-level trait-trait associations between trophic levels. Specifically: a) Tall palms with medium-sized fruits (p3) are strongly associated with small to medium sized mammals that consume moderate to high amounts of fruit (m1). b) Acaulescent or small-stemmed palms with small to large fruits (p6) correlate strongly with small mammals with relatively low frugivory (m3). Finally, c) medium-sized to large sized frugivorous mammals with moderate to low frugivory levels (m6) interact with moderately tall erect palms with medium-sized fruits (p5). Our approach resulted in 1287 grid cells across the Neotropics, each containing seven measures of functional richness (FR) for palms and seven measures for mammals, corresponding to each interaction guild within the two trophic levels. Across all grid cells, mammalian frugivore guild m3 dominated (54 % ± 9), comprising 341 species of relatively small-bodied frugivores (mean 4.75 ± 1.33 kg) with low fruit intake (21.7 ± 7.2 %) and predominantly nocturnal habits (84 % nocturnal, 22 % diurnal). Its counterpart in palms (p3) was much smaller (just 15 species) but consisted of the tallest stems (3.45 ± 0.26 m) bearing the largest fruits (1.69 ± 0.40 cm). Mammalian frugivore guild m6 (15 % ± 5) included 15 of the largest frugivores (9.73 ± 0.73 kg) with moderate frugivory (20 ± 12.5 %) and mixed activity patterns (40 % diurnal, 73 % nocturnal, 47 % crepuscular); its counterpart palm assemblage (p6) (137 species) was dominated by short‐stemmed, large‐fruited forms (1.18 ± 0.96 m; 1.39 ± 0.57 cm) with a high proportion of acaulescent growth (36 %). Smaller frugivorous mammal assemblages in mammalian guilds m1 (12 % ± 4) and m7 (10 % ± 4) comprised 66 and 33 species, respectively: m1 species were moderately sized (6.47 ± 1.68 kg), highly frugivorous (64.1 ± 11.6 %) and evenly split between diurnal and nocturnal, whereas m7 species were larger (8.15 ± 0.52 kg), moderately frugivorous (36.4 ± 4.9 %) and mostly diurnal (79 %). Guild m4 (9 % ± 5) included 33 small (3.91 ± 0.66 kg), highly frugivorous mammals (48.5 ± 3.6 %) that were almost exclusively nocturnal (97 %). Among palms, guild p5 dominated (46 % ± 15) with 265 low‐stature (understory), erect stemmed palms (1.95 ± 0.63 m; 0.90 ± 0.31 cm fruits; 1 % acaulescent), followed by guild p6 (34 % ± 18) and guild p3 (10 % ± 7), while guilds p1, p4 and p7 had no palm representatives (Table S4, Figure S2a, S2b).

### Assessing the influence of climate on functional richness

Overall, the functional richness of palm interaction guilds showed a weak association with climatic gradients (Table S1, Figure S3). Only mean annual temperature had a significant positive effect (estimate = 4.07e−04, P = 0.0027), while precipitation (P = 0.23), temperature seasonality (P = 0.79), and precipitation seasonality (P = 0.50) showed no significant effects (Table S1). In contrast, the functional richness of mammal interaction guilds was significantly and positively related to temperature (estimate = 4.42e−04, P \< 0.001), but unrelated to precipitation (P = 0.15), temperature seasonality (P = 0.63), or precipitation seasonality (P = 0.13) (Table S2). However, when using smoothed terms for separating the effects of climate across individual interaction guilds, we found significant correlations (Table S1). Mean annual temperature had a significant nonlinear effect for palm guild p3 (F = 6.30, P = 0.014), whose functional richness declined with increasing temperature. Several other guilds (m1, m4, m5, m7) showed marginally significant temperature effects (P \< 0.1), but responses were weak and inconsistent in direction. Significant precipitation effects were only observed for palm guild p6 (F = 2.95, P = 0.0067), which showed a positive association, indicating increased functional richness of short-stemmed or acaulescent palms in wetter environments. No other guilds showed a statistically significant relationship with precipitation. For precipitation seasonality, statistically significant responses were observed for palm guild p5 (F = 8.37, P = 0.0070) and IG-6 (F = 5.62, P = 0.0011). These had opposite directions: p5 functional richness decreased with higher seasonality, while it increased for p6. No guild showed statistically significant effects of temperature seasonality (P \> 0.35 for all) (Figure 4a, Figure S1, Table S1). At the opposite trophic level, mean annual temperature emerged as the only statistically significant climatic driver, with significant nonlinear effects for mammalian guilds m1 (F = 6.66, P = 0.011), m2 (F = 6.91, P = 0.010), m4 (F = 11.21, P = 0.001), m5 (F = 6.91, P = 0.010), m6 (F = 10.33, P = 0.0016), and m7 (F = 9.02, P = 0.0032). Visual inspection of the partial effect plots revealed that all statistically significant guilds exhibited declining functional richness with increasing temperature, suggesting a negative nonlinear response within guilds (Figure 4a, Figure S1, Table S2).

![**Figure 4. Relationships between environmental gradients and functional richness, functional trophic asymmetry of mutualistic palm-mammal frugivore networks in the Neotropics.** Figure panels a and b illustrate how functional richness (left) and functional trophic asymmetry (FTA; right) of palm–mammal frugivore guilds vary along two climatic gradients. In panel a (top), mammal guilds (red lines) show relatively flat functional richness across mean annual temperature, whereas palm guilds (green lines) diverge: guild IG-3p increases sharply toward warmer areas while IG-5m and IG-6m remain constant. In the lower‐left panel, functional richness declines with increasing precipitation seasonality in mammal guilds IG-3m, IG-4m, and IG-7m, but palm guild IG-6p increases, reflecting a shift toward more functionally diverse short‐stemmed species under seasonal rainfall. Panel b (top) reveals that FTA of certain mammal‐only guilds (e.g. m3p3 in yellow and m3p6 in red) steadily rises with temperature, while other guilds remain unchanged. The bottom‐right panel shows a more complex mosaic under precipitation seasonality, with some guilds (m5p6, m4p6) displaying stronger asymmetry in highly seasonal environments and others (m1p6, m7p6) remaining stable. Together, these plots underscore that palms and mammalian frugivores respond idiosyncratically to climatic gradients, and that the balance of their trait distributions can be either buffered or exacerbated by temperature and rainfall seasonality depending on the interaction guild.](images/clipboard-559875527.png)

### Assessing the influence of climate on functional trophic asymmetry

The distribution of grid cell measures of functional trophic asymmetry varied significantly among interaction guilds in palm–mammal frugivore networks (Figures 4b). The highest asymmetries were observed in guild combinations (m = mammals, p = palms) such as m3p7, m3p4, and m3p1, indicating strong differences between functional traits of palms and associated mammalian frugivores. In contrast, guilds like m7p2, m4p3, and m7p3 showed markedly lower functional trophic asymmetry values (Figure S4). The weighted sum of guild-level Functional Trophic Asymmetry (FTAsum), which summarizes FTA variation across all interaction guilds, averaged 0.91 across all grid cells (SD = 0.10), with values ranging from 0.67 to 1.28 (Figure S6). FTAsum was positively correlated with mean annual temperature and precipitation seasonality. In other words, greater differences in the functional richness between trophic levels peaked in warm and regions with seasonal rainfall (Figure 5, Figure 6). Functional trophic asymmetry was positively related to mean annual temperature across all interaction guilds (F = 5.97, P = 0.01). However, no consistent relationships were found with precipitation (F = 0.28, P = 0.13), precipitation seasonality (F = 0.30, P = 0.59), or temperature seasonality (F = 0.19, P = 0.67) (Table 1).

![**Figure 5: Spatial variation of functional trophic asymmetry (FTAsum) and network-level specialization (H2′) across the Neotropics.** Panels a & b shows the spatial variation of climatic predictors, i.e. (a) mean annual temperature, and (b) precipitation seasonality. Panel (c) shows FTAsum, which quantifies the disparity in functional richness between interacting trophic levels, with higher values (red) indicating greater asymmetry. Panel (d) displays H2′, a standardized measure of interaction specialization, where higher values (red) represent more specialized networks. The figure illustrates how broad‐scale climate gradients shape overall functional‐trait mismatch (FTAsum) and network specialisation (H2′) in Neotropical palm–mammal frugivore assemblages. In panel a, mean annual temperature (MAT) peaks above 28 °C in lowland Amazonia and Atlantic Forest and falls below 10 °C in the high Andes, while precipitation seasonality (PS) is lowest (\< 30 mm) in the ever-wet Amazon and exceeds 120 mm in seasonally dry forests.](images/clipboard-2626677919.png)

![**Figure 6: The variation of functional trophic asymmetry and network specialization for palm-mammal frugivore seed dispersal networks across climatic gradients in the Neotropics.** Panels show partial effects (±95 % CI) of a generalized additive model: FTAsum increases steadily with mean annual temperature, indicating greater functional trophic asymmetry in warmer areas, but is effectively flat across the precipitation seasonality gradient. By contrast, H2′ remains low and invariant up to \~18 °C before rising at higher mean annual temperature, increasing monotonically with precipitation seasonality, indicating more specialised interactions under pronounced rainfall seasonality.](images/clipboard-2486706043.png)

### Assessing how interaction strength mediates the influence of climate on functional trophic asymmetry

The strength of interactions between guilds of consumers and producers had an overall positive but weak effect on functional trophic asymmetry (Figure 4b, Figure S7, Table S3). Moreover, we did not find strong evidence supporting that the relationship between functional trophic asymmetry and climate depends on the strength of interaction between interaction guilds (F = 746, P \> 0.05) (Figure S7, Table 1). The interaction term between guild interaction strength and precipitation seasonality had a marginally significant positive effect on functional trophic asymmetry (F = 0.14, P = 0.051) (Figure S8, Table 1). Assessing the relationship between functional trophic asymmetry and network specialization We found that palm–mammal frugivore networks exhibited moderate to occasionally high levels of trophic specialization, with H2′ values ranging from 0.08 to 0.79 (median = 0.21), where 0 means that species have no preference for their interaction partners and 1 represents specialized networks where each species interacts only with a specific subset of interaction partners (Figure S9). H2′ was positively associated with the functional trophic asymmetry index (FTAsum) (β = 0.047, P = 0.002; Table S3), indicating that networks with greater asymmetry between consumer and resource traits tended to be more specialized (Figure 6, Figure S9). In addition, H2′ was significantly related to mean annual temperature (F = 3.39, P = 0.005) and precipitation seasonality (F = 3.13, P = 0.003), suggesting that environmental gradients shape patterns of trophic specialization (Figure 6). In contrast, H2′ showed no significant association with temperature seasonality or total annual precipitation (P \> 0.2). The full model explained 4.5% of the H2’ deviance.


::: callout 
#### Table 1: Summary of parametric and smooth term coefficients for the model predicting FTA responses to climate.

The only parametric term inferred was for the intercept, and we provide its coefficient estimate, standard error, t-value, and p-value. Smooth terms are presented for each non-linear predictor and its interactions. Statistically significant p-values are highlighted in bold. The term, int_str refers to the “interaction strength” between a guild pair combination.

```{r}
#| message: false
#| warning: false
#| echo: false

fta_table <- tribble(
  ~Predictor, ~Estimate, ~CI, ~p,
  "(Intercept)", "0.21", "0.21 – 0.21", "<0.001",
  "s(Mean Annual Temperature)", NA, NA, "0.007",
  "s(Mean Annual Temperature × Interaction Strength)", NA, NA, "0.383",
  "s(Total Annual Precipitation)", NA, NA, "0.636",
  "s(Total Annual Precipitation × Interaction Strength)", NA, NA, "0.030",
  "s(Temperature Seasonality)", NA, NA, "0.763",
  "s(Temperature Seasonality × Interaction Strength)", NA, NA, "0.421",
  "s(Precipitation Seasonality)", NA, NA, "0.776",
  "s(Precipitation Seasonality × Interaction Strength)", NA, NA, "0.289",
  "s(Interaction Strength)", NA, NA, "<0.001"
) %>%
  mutate(
    p_num = if_else(
      str_detect(p, "<"),
      0.001,
      as.numeric(p)
    )
  )


fta_table %>%
  gt() %>%
  tab_header(
    title = md("**Functional Trophic Asymmetry (FTA)**"),
    subtitle = md("Generalized Additive Model results")
  ) %>%
  cols_label(
    Predictor = "Predictor",
    Estimate  = "Estimate",
    CI        = "95% CI",
    p         = html("p-value")
  ) %>%
  fmt_missing(
    columns = c("Estimate", "CI"),
    missing_text = ""
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      columns = p,
      rows = p_num < 0.05
    )
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(
      rows = Predictor == "(Intercept)"
    )
  ) %>%
  tab_source_note(
    source_note = md("Observations = 53,406 &nbsp;&nbsp;|&nbsp;&nbsp; R² = 0.200")
  ) %>%
  cols_hide(p_num) %>%
  opt_table_outline() %>%
  opt_all_caps()
```
:::

::: callout 
#### Table 2 Summary of parametric and smooth term coefficients for the model predicting H2’ responses to environmental and FTA variations.

The only parametric term inferred was for the intercept, and we provide its coefficient estimate, standard error, t-value, and p-value. Significant p-values are highlighted in bold and suggest predictors with a meaningful contribution to the model.

```{r}
#| echo: false
#| message: false
#| warning: false

ns_table <- tribble(
  ~Predictor, ~Estimate, ~CI, ~p,
  "(Intercept)", "0.16", "0.16 – 0.17", "<0.001",
  "FTAsum", NA, NA, "<0.001",
  "s(Mean Annual Temperature)", NA, NA, "0.040",
  "s(Precipitation Seasonality)", NA, NA, "0.029",
  "s(Temperature Seasonality)", NA, NA, "0.391",
  "s(Total Annual Precipitation)", NA, NA, "0.482"
) %>%
  mutate(
    p_num = if_else(str_detect(p, "<"), 0.001, as.numeric(p))
  )

ns_gt <- ns_table %>%
  gt() %>%
  tab_header(
    title = md("**Network Specialization**"),
    subtitle = md("Model results")
  ) %>%
  cols_label(
    Predictor = "Predictors",
    Estimate  = "Estimates",
    CI        = "CI",
    p         = html("p")
  ) %>%
  fmt_missing(
    columns = c("Estimate", "CI"),
    missing_text = ""
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = p, rows = p_num < 0.05)
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = Predictor == "(Intercept)")
  ) %>%
  tab_source_note(
    source_note = md("Observations = 1,242 | R² = 0.160")
  ) %>%
  cols_hide(p_num) %>%
  opt_table_outline()

ns_gt
```
:::

## Discussion

Our study reveals that the functional traits relevant to trophic interactions are subject to different relative strengths of environmental filtering among trophic levels, and over space in the Neotropics. In palm-mammal frugivore assemblages, these differences (a.k.a. functional trophic asymmetry) reach their maximum in areas with warm climates and highly seasonal precipitations. Furthermore, multitrophic species assemblages with high FTA paradoxically exhibited a high level of specialization in their trophic interaction networks. Taken together, our results suggest that distinct community assembly processes operate across trophic levels, reinforcing the idea that network assembly emerges from a dynamic interplay of bottom-up (producer driven) and top-down (consumer driven) processes [@moretti2009combining]; @marjakangas2022trait; @schleuning2023animal\] , whose respective contributions are dependent on the environmental and biotic conditions.

### Contrasting trait-mediated responses to environmental variation across trophic levels

Consumer and producer assemblages are known to differ in their functional response to climatic gradients [@moretti2009combining; @dehling2014functional; @albrecht2018; @song2024trophic ]. In our Neotropical palm-frugivore study system, changes in functional trophic asymmetry along the temperature gradient were primarily driven by consumers (mammalian frugivores), reflecting an increased representation of frugivores with a low proportion of fruit in their diet in warmer are-as. Along with the precipitation seasonality gradient, shifts in functional trophic asymmetry were producer-driven, linked to a decline in short-stature (understory) palms with low to medium fruit sizes and a concurrent rise in short-stemmed (acaulescent) palms bearing medium to large fruits (Figure 4). For mammalian frugivores, functional richness generally responded to the tempera-ture gradient, except for the guild representing small-bodied frugivores (IG-3), which exhibited relative stability across the gradient. These small-bodied frugivores can be primarily seed preda-tors rather than effective seed dispersers. As such, they may also be contributing more to antag-onistic than mutualistic outcomes in palm–frugivore interactions [@marques2022mutualism] (Table S1). Palm responses are more tightly coupled to variation in precipitation seasonality and total annual precipitation, with both the shape and strength of these responses varying among guilds. Notably, there is a functional shift towards warmer regions with high precipitation seasonality, where palms with relatively small fruits (p5) are replaced by palms bearing relatively large fruits (p6). This pattern is consistent with previous studies showing that palm assemblages in regions with strong climatic seasonality and long dry periods tend to have larger fruit sizes in addition to a reduced species richness, likely an adaptive response to historical climatic fluctuations during the Quaternary [@kissling2012towards; @goldel2015geographical].

### Top-down and bottom-up variations in functional trophic asymmetry are mediated by climate

Functional trophic asymmetry varied geographically, peaking in warm regions with high seasonal precipitation. Overall, average annual temperature has a small but positive effect on functional trophic asymmetry. The effect on functional trophic asymmetry was explained by larger temperature-driven shifts in the functional richness of mammals than in that of palms. These results support the view that the assembly of palm-frugivore networks inhabiting the warmest regions of the Neotropics is predominantly top-down (i.e., consumer-driven). In contrast, the detected influence of precipitation seasonality on the functional richness of palm assemblages supports the view that the assembly of palm- mammalian sfrugivore networks in areas with highly seasonal precipitations is mostly bottom-up (i.e., producer-driven) [@schleuning2016ecological; @albrecht2018; @marjakangas2022trait; @dehling2020similar].

Interaction strength, that is, the intensity of the effect that one species has on another within an ecological network, varies strongly across palms and mammalian frugivore guilds. Moreover, variations in the strength of interaction among palm-frugivore guilds mediate the response of functional trophic asymmetry to precipitation seasonality. Specifically, the level of functional trophic asymmetry in strongly interacting guilds shows a more significant and positive relationship with precipitation seasonality. This means, for example, that palm species from guilds with more specialized seed dispersal are more common while palm species from guilds with a generalist seed dispersal strategy become rarer towards regions where rainfall is seasonal (Figure 4b, Figure 6). As precipitation becomes more seasonal, Neotropical palms are represented by species with larger fruits [@kissling2012towards] and mammalian frugivores with a broad diet become scarcer. For instance, the biomass of frugivorous primates is generally lower in seasonal environments with marked fruiting periods [@hanya2011seasonality]. Our results align with findings of bat-plant mutualistic pollination networks, where interacting species exhibit lower niche overlaps in highly seasonal environments, and with those of avian-frugivory networks where weaker trait matching is found towards the tropics [@schleuning2012specialization; @huang2025weaker]. They also contrast with others that report stronger trait-matching and interaction strength towards the tropics in plant-hummingbird pollination networks [@sonne2020ecological] and avian-palm frugivory networks [@mcfadden2022global]. While our results are directly relevant to palm-mammal seed dispersal interactions, the diversity of process across mutualistic systems highlights the context dependency of trait–interaction-climate relationships. As such, generalizations about the functional ecology of mutualistic networks should pay attention to partner identity and interaction type.

### Functional trophic asymmetry as a multitrophic indicator of ecological processes and their response to global changes

The concept of functional trophic asymmetry offers a valuable framework to examine how multitrophic ecological communities respond to shifting environmental conditions. Functional trophic asymmetry captures how environmental filtering differentially affects producers and consumers, revealing the contributions of bottom-up and top-down processes in shaping ecological network structure. While past studies have largely focused on single-trophic responses to climate or trait matching within networks, the framework we present here combines network assembly modeling with trait-based analyses, facilitating the generation of multiscale synthetic species interaction data, as well as ecological interpretations of broad-scale multitrophic community assembly pat-terns within a single framework. In addition to climate change, habitat destruction, defaunation, and other human-induced landscape modifications can independently alter resource and consumer functional diversity, thereby affecting functional trophic asymmetry [@guevara2023asymmetric; @bello2023analyzing; @guerra2025human]. Although differences in the extent of landscape modification were not directly investigated in this study, previous research in plant-hummingbird mutualisms (i.e. pollination) shows that deforestation has a greater impact on the diversity and redundancy of functional roles in plants than in their pollinators, suggesting a de-cline in plant functional roles as forests are lost [@tinoco2018land; @guevara2023asymmetric].

Future work may focus on integrating additional information such as geographic variation in population density, land-use change, or spatial movement data (e.g., GPS tracking of frugivores) that would improve realism and generalization of our modelling approach to generate synthetic ecological networks [@borah2022studying; @beumer2025movetraits; @cousens2010towards]. For instance, the assumption of higher species abundances toward the center of geographic ranges could be challenged, as spatial patterns in species abundance might depend on how geographic ranges are aligned with background environmental gradients [@schmidt2022population]. Moreover, the role and importance of some individual consumer species in plant-animal seed dispersal networks (e.g. species’ centrality) might vary with the distance to their climatic niche centroid, suggesting the existence of areas where ecological and evolutionary dynamics might be elevated [@moulatlet2025general]. New elements such as these could be integrated into the analytical pipelines once sufficient relevant data becomes digitally available.

## Conclusion

As climate change accelerates, rising temperatures and altered precipitation regimes are likely to increase functional trophic asymmetry in multitrophic palm-frugivore communities and increase specialization in palm seed dispersal networks. A likely consequence is also the loss of ecosystem function, in particular the zoochorous dispersal of palm propagules between habitat patches, which has been shown to decrease with increased specialization and reduced plant functional diversity [@landim2025functional]. Such structural changes may also arise in other mutualistic net-work systems supporting key ecosystem functions, including pollination, myrmecochory, and plant-soil microbiome networks [@schleuning2014ecological; @schleuning2020trait]. These could similarly experience reduced functional connectivity and lower resilience due to increased specialization and declining functional diversity under climate change [@donoso2017complementary; @caradonna2021seeing; @cruz2023mutualisms]. However, future research should explicitly test these predictions across different ecological systems to better understand the generality and scale of these impacts [@landim2025functional; @landim2025functional; @rabeau2025projected; @acosta2023climate]. As species distributions shift in response to climate change [@Aizen2012; @ValienteBanuet2015; @Bartley2019; @Hurtado2024], our work showcases the potential role of functional trophic asymmetry as a key framework to advance our ability to predict shifts in ecosystem function [@Montoya2010; @Bello2023; @Heilmeier2019] with incomplete multi-trophic data on species interactions and traits.
